---
title: "PS6"
output: html_document
date: "2024-05-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.1 
```{r}
scripps <- read.delim("scripps-co2.csv")
head(scripps)
```
```{r}
library(dplyr)
scripps2 <- scripps %>%
  select(year, month, co2)
head(scripps2)
```

## 1.2 
Each row in this data set represents a monthly CO2 measurement at a different date and time.

## 1.3 
There are no missing values in this data set. 

## 1.1.4
The data ranges all the way from 1958 to 2023

## 1.1.5
```{r}
library(ggplot2)
scripps2 %>% 
  filter(as.numeric(co2) > 0) %>%
  mutate(months = as.Date(paste(year, month, "01", sep = "-"))) %>% 
  ggplot(aes(months, co2)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 0.6) +
  scale_x_date(date_labels = "%Y", date_breaks = "5 year") +
  labs(title = "CO2 Concentration in Atmosphere vs Years", x = "Year", y = "CO2")
```

## 1.1.6
Each year the, CO2 level has been increasing from 1958 till 2023.

## 1.2.1
```{r}
hadcrut <- read.delim("hadcrut-5.0.2.0-annual.csv")
head(hadcrut)
```
```{r}
library(dplyr)
hadcrut2 <- hadcrut %>%
  select(year = Time, Anomaly = Anomaly..deg.C.) %>%
  rename(temperature_anomaly = Anomaly)
head(hadcrut2)
```

## 1.2.2
Each line represents each years temperature anomaly

## 1.2.3
1850 to 2023

## 1.2.4
```{r}
library(dplyr)
preIndustrial <- hadcrut2 %>%
  filter(year <= 1900) %>%
  summarise(preIndustrialAvg = mean(temperature_anomaly, na.rm = TRUE))
preIndustrial
```
```{r}
hadcrut2 <- hadcrut2 %>%
  mutate(anomaly_over_preindustrial = temperature_anomaly - preIndustrial$preIndustrial)
head(hadcrut2)
```

## 1.2.5
It means that the average global temperature during the pre industrial period is lower than the reference period average.

## 1.2.6
```{r}
library(dplyr)
pre_industrial_data <- hadcrut2 %>% filter(year <= 1900)
pre_industrial_baseline <- mean(pre_industrial_data$temperature_anomaly, na.rm = TRUE)
cat("Pre-Industrial Baseline (°C):", pre_industrial_baseline, "\n")
hadcrut2 <- hadcrut2 %>%
  mutate(Temp_Anomaly_Pre_Industrial = temperature_anomaly - pre_industrial_baseline)

head(hadcrut)
```

## 1.2.7
```{r}
library(dplyr)
library(ggplot2)

hadcrut2 <- hadcrut2 %>%
  mutate(Temp_Anomaly_Pre_Industrial = temperature_anomaly - pre_industrial_baseline)

# Plot the data
ggplot(hadcrut2, aes(x = year, y = Temp_Anomaly_Pre_Industrial)) +
  geom_line(color = "blue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "Global Surface Temperature Change Over Time",
       x = "Year",
       y = "Temperature Anomaly (°C) above Pre-Industrial Baseline") +
  theme_minimal()
```

## 1.2.8
Over 170 years, the global temperature has slowly increase from 1850 to around 1960. After that the temperature rises higher.

## 1.3.1
I would use the variable year to merge co2 data with global temperature.

## 1.3.2
By creating a new variable for the data set that represents the year

## 1.3.3
```{r}
library(dplyr)
library(ggplot2)
CO2_Year <- scripps2 %>% 
  group_by(year) %>%
  summarize(avg_co2_yearly = mean(co2, na.rm = TRUE)) 
 merger <- left_join(hadcrut2, CO2_Year)
```
I did a left join as it allows us to combine using the year variable.

## 1.3.4
```{r}
library(dplyr)
library(ggplot2)
merger %>%
  filter(avg_co2_yearly >= 300) %>% 
  ggplot(aes(avg_co2_yearly, anomaly_over_preindustrial, col = year)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(color = "black") +
  geom_hline(yintercept = 1.5, color = "black", size = 1) +
  scale_alpha_continuous(name = "decades") +
  scale_color_gradient(low = "green", high = "red")
```

## 2.1
```{r}
library(dplyr)
UAH <- read.delim("UAH-lower-troposphere-wide_2023-11-07.csv") %>% 
  select(year,month,globe)
head(UAH)
```

## 2.2
```{r}
library(dplyr)
UAHaverages <- UAH %>%
  group_by(year) %>%
  summarize(lowerTroposphere = mean(globe))
UAHaverages %>% 
  head()
```

## 2.3
```{r}
library(dplyr)
hadcrut2$surface <- hadcrut2$anomaly
UAHHadcrut2 <- left_join(UAHaverages, hadcrut2, by = "year") %>% 
  select(!c(anomaly_over_preindustrial,temperature_anomaly))
UAHHadcrut2 %>% 
  head()
```

## 2.4
Data set is in wide form as seen when opening up the data set from the environment. 

## 2.5
```{r}
library(dplyr)
library(magrittr)
library(tidyverse)
LongUAHHadcrut <- UAHHadcrut2 %>% 
  pivot_longer(!c(year),names_to="type",values_to="anomaly")
LongUAHHadcrut %>% 
  head()
```

## 2.6
```{r}
library(dplyr)
library(ggplot2)
UAHHadcrut2 %>% 
  ggplot()+
  geom_point(aes(year,lowerTroposphere),
             color="orange")+
  geom_smooth(aes(year,lowerTroposphere),
                  method="lm",se=FALSE,color="orange")+
  geom_point(aes(year,surface),
             color="purple")+
  geom_smooth(aes(year,surface),
                  method="lm",se=FALSE,color="purple")+
    labs(
    x="Year",
    y="Temperature Anomaly",
  )
```

## 3.1
```{r}
baby <- read.delim("babynames.csv")
head(baby)
```

```{r}
library(dplyr)
filtered_baby_data <- baby %>%
  filter(name %in% c("Mary", "Emma"))
head(filtered_baby_data)
```

## 3.2
```{r}
library(dplyr)
library(magrittr)
library(tidyverse)
wide_baby_data <- filtered_baby_data %>%
  pivot_wider(names_from = name, values_from = n)
head(wide_baby_data)
```

## 3.3
```{r}
library(dplyr)
library(ggplot2)

filtered_baby_data$interval <- cut(filtered_baby_data$year, breaks = seq(1900, 2020, by = 20))
ggplot(filtered_baby_data, aes(x = year, y = n, color = interval, shape = name)) +
  geom_line(aes(group = name)) +
  geom_point() +
  scale_color_discrete(name = "20-Year Interval") +
  labs(title = "Popularity Trajectory of Names Over 20-Year Intervals",
       x = "Year",
       y = "Count",
       shape = "Name") +
  theme_minimal()
```

## 3.4 
Before 2000, Mary would easily be a much more popular named compared to Emma. However, suddenly, in the 2000s, Emma became much more popular of a name compared to Mary.

